
# Plotting

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(fig.width = 5)
knitr::opts_chunk$set(fig.height = 6)
library(SeaVal)
```

The function `ggplot_dt` takes a data table containing two columns named `lon` and `lat` that should specify a regular longitude-latitude grid, as well as some data to use for coloring the map. An easy example is the following

```{r}
data("chirps_monthly")

dt = copy(chirps_monthly) # to manipulate the data table: chirps_monthly has locked binding

dt2020 = dt[year == 2020 & month == 10] # reduce the observed precipitation data to a single time slice, namely October 2020
print(dt2020) 
ggplot_dt(dt2020,'prec')
```

As we can see, the color scale here makes no sense (blue meaning no precipitation) and we'll talk about that in a second. But let's start at the base functionality. `ggplot_dt` requires two arguments, the first one being a data table containing the data, and the second one is the name of the column that contains the data you want to plot. This defaults to the third column in the data table (often your data table will start with lon, lat, and the third column is what you want to plot). So in the example above, `ggplot_dt(dt)` would have led to the same result, because `'obs'` is the third column in `dt`. The plotting window is determined by the data. If you have data covering the entire earth, the entire earth would be plotted. As a consequence, we can restrict the plotted region by subsetting the data table:
```{r}
dt_sub  = dt2020[lon %between% c(28,43) & lat %between% c(-12,1)] # a region containing Tanzania
ggplot_dt(dt_sub)
```

The function has further optional arguments (also recall that you can access the function documentation summarizing all of this by typing `?ggplot_dt`):
```{r}
ggplot_dt(dt_sub,'prec',
          mn = 'October 2020 precipitation', # add a title to the plot
          rr = c(1,10), # fix the limits of the color scale
          name = 'mm/day') # change the legend label
```

In this example we set the lower limit of the color scale to 1 and the upper limit to 10. Note that the plot truncates the data at the ends of the color scale, so every pixel with precipitation of  below 1mm/day is now shown in the same blue color. Setting the range of the color scale is particularly useful to force symmetry around 0, e.g. when plotting correlations or anomalies:
```{r}
dt[,clim := mean(prec), by = .(lon,lat,month)] # add a climatology column
dt[,ano := prec - clim] # add anomaly as another column
print(dt)
ggplot_dt(dt[month == 10 & year == 2020], 'ano', rr = c(-3,3))

```

Now, in this plot positive rainfall anomalies are shown red while negative anomalies are blue, which is very counterintuitive. The function allows us to specify the three used colors by name with the arguments `low`,`mid`, and `high`. So here's an anomaly plot looking a bit nicer:
```{r}
ggplot_dt(dt[month == 10 & year == 2020], 'ano', 
          mn = 'October 2020 rainfall anomaly',
          rr = c(-3,3),
          low = 'red', mid = 'white', high = 'darkgreen',
          name = 'mm/day')
```

Note that we set the range argument to `c(-3,3)`. Fixing the range makes mostly sense when the range is known (e.g. for correlation plots), or when you want to compare several plots (e.g. for comparing mean square error of different NWP models, all plots should have the same range). If we leave the range argument `rr` free, the range is determined from the data. However, when we do this for our anomaly plot this has an undesired sideeffect:
```{r}
ggplot_dt(dt[month == 10 & year == 2020], 'ano', 
          mn = 'October 2020 rainfall anomaly',
          low = 'red', mid = 'white', high = 'darkgreen',
          name = 'mm/day')
```

The color scale is no longer centered (white) at 0, but in the center of the (now asymetric) range. As a consequence, all gridpoints with anomaly 0 are shown in a light red. To fix this we can use the `midpoint` argument:
```{r}
ggplot_dt(dt[month == 10 & year == 2020], 'ano', 
          mn = 'October 2020 rainfall anomaly',
          low = 'red', mid = 'white', high = 'darkgreen',
          midpoint = 0,
          name = 'mm/day')
```

Another, maybe surprising, use for the `midpoint` argument is that we can generate plots with a colorscale with only two colors. For example, going back to plotting the observed rainfall we can do the following:
```{r}
ggplot_dt(dt[month == 10 & year == 2020], 'prec', 
          mn = 'October 2020 rainfall',
          mid = 'white', high = 'blue',
          midpoint = 0,
          name = 'mm/day')
```

What happens here is that we set the midpoint to 0, which is the minimum of our data (since observed rainfall is never below 0). Consequently the second half of the colorscale extending below 0 is ignored.

Finally, the function allows to discretize the color scale. To this end we need to set the argument `discrete_cs` to `TRUE`. We can then control the breaks of the discrete colorscale by either of the arguments `binwidth`, `n.breaks`, or `breaks` (the latter takes a vector containing all breakpoints). To revisit the anomaly plot from above:
```{r}
ggplot_dt(dt[month == 10 & year == 2020], 'ano', 
          mn = 'October 2020 rainfall anomaly',
          discrete_cs = TRUE, binwidth = 2,
          low = 'red', mid = 'white', high = 'darkgreen',
          midpoint = 0,
          name = 'mm/day')
```


One final remark: Often you will deal with data tables that contain spatio-temporal data. It is then important to subselect the particular timeslice you want to plot before plotting (as we did above when we plotted October 2020). The function `ggplot_dt` by default tries to select the first timeslice of tempo-spatial data. This is often not what you want (which is why you should subselect first), but it's frequently handy for quick data diagnostics, which is the reason for this default behavior. Here an example:
```{r}
print(chirps_monthly) # a data table with multiple months and years and locations, so spatio-temporal data
ggplot_dt(chirps_monthly) # generates a plot of the precipitation of October 1981 (first timeslice), for a first quick impression how your data looks
```

## Plotting values for selected countries

Above, we have already seen an option how to restrict a plot to a particular country: by manually subsetting the data to a rectangle of longitudes and latitudes containing that specific country. This is of course quite tedious, and to make our lives easier we can use the `restrict_to_country`-function that takes a data table and a country name, and subsets the data table to only contain gridpoints in the specified country. Currently, the function accepts the following country names: Burundi, Eritrea, Ethiopia, Kenya, Rwanda, Somalia, South Sudan, Sudan, Tansania, Uganda.
```{r}
dt_new = restrict_to_country(dt[month == 10 & year == 2020],'Kenya')
print(dt_new)
ggplot_dt(dt_new, 'ano', 
          mn = 'October 2020 rainfall anomaly',
          discrete_cs = TRUE, binwidth = 2,
          low = 'red', mid = 'white', high = 'darkgreen',
          midpoint = 0,
          name = 'mm/day')

```

As we can see, the function restricts the data to all gridcells for which the centerpoint lies within the specified country. This is useful, for example, for calculating mean scores for the specified country. However, it is not optimal for plotting since every grid cell past the border is censored even though the original data table contained values there. To this end, the `restrict_to_country` function has a `rectangle`-argument that you can set to `TRUE` for plotting:
```{r}
dt_new = restrict_to_country(dt[month == 10 & year == 2020],'Kenya', rectangle = TRUE)
ggplot_dt(dt_new, 'ano', 
          mn = 'October 2020 rainfall anomaly',
          discrete_cs = TRUE, binwidth = 2,
          low = 'red', mid = 'white', high = 'darkgreen',
          midpoint = 0,
          name = 'mm/day')

```

Instead of a single country name you can also pass multiple country names in a vector to the function. Moreover, when you use `rectangle = TRUE`, you can specify a tolerance `tol` in order to widen the plotting window:
```{r}
dt_new = restrict_to_country(dt[month == 10 & year == 2020],
                             c('Kenya','Tanzania'),
                             rectangle = TRUE,tol = 2)
ggplot_dt(dt_new, 'ano', 
          mn = 'October 2020 rainfall anomaly',
          discrete_cs = TRUE, binwidth = 2,
          low = 'red', mid = 'white', high = 'darkgreen',
          midpoint = 0,
          name = 'mm/day')

```

The `tol = 2` argument means that the function will include a buffer zone of 2 degrees lon/lat outside the specified countries (i.e. 4 gridpoints to each side). Note that the buffer to the south of Tanzania is smaller, because the original data table `dt` does not contain any data further south.


## More plotting options

The function `ggplot_dt` is, as its name suggests, based on the package `ggplot2`. In `ggplot2`, plots are composed out of multiple layers, allowing for successive adding of layers. This can help us to generate highly customized plots. As an example, let's revisit the anomaly plot from above and add the location of Nairobi an Addis Abbaba to it:
```{r}
library(ggplot2)

# get locations as data table:
loc = data.table(name = c('Addis Abbaba','Nairobi'),lon = c(38.77,36.84),lat = c(9,-1.28))
print(loc)

pp = ggplot_dt(dt[month == 10 & year == 2020], 'ano', 
               mn = 'October 2020 rainfall anomaly',
               low = 'red', mid = 'white', high = 'darkgreen',
               midpoint = 0,
               name = 'mm/day') + 
     geom_point(data = loc,mapping = aes(x = lon,y = lat)) + 
     geom_text(data = loc,mapping = aes(x = lon,y = lat,label = name),vjust = 1.5)

print(pp)

```

Here, we added two layers to the original plot, the first one being the `geom_point`-layer that creates the two points at the locations of the cities, and the second being the `geom_text`-layer that labels the points by the city names. `ggplot2` is a widely used package and there is a large variety of tutorials and books out there that can help you getting familiar with the syntax.

A frequently required operation is the changing of the font sizes of title and labels. The easiest way to do this is the command
```{r}
theme_set(theme_bw(base_size = 16)) # defaults to 12
print(pp)
```

If we want to print the plot to an external file, all we have to do is
```{r, eval = FALSE, echo = TRUE}
pdf(file = '<path to file and filename>.pdf', width = ...,height = ...)
  print(pp)
dev.off()
```
This prints a .pdf file, but you can print .png and some other file formats similarly, see `?Devices` for an overview.

We can also use ggplots adding-layer-syntax to overwrite existing layers, for example if we want a fully customized colorscale:
```{r, fig.width = 4}
library(viridis) # the viridis package contains some nice color scales
pp_new = pp + scale_fill_viridis(name = 'my_color_scale',
                                 breaks = seq(-5,5,by = 2),
                                 guide = guide_colorbar(title = 'my personal color scale',
                                                        title.position = 'top',
                                                        barwidth = 20,
                                                        direction = 'horizontal')) +
              theme(legend.position = 'bottom')
print(pp_new)
```

