---
title: "Validation of Forecasts with SeaVal"
output: html_document
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(fig.width = 6)
knitr::opts_chunk$set(fig.height = 6)

```

This sheet gives a short introduction how to use the `R`-package 'SeaVal' for seasonal validation. This package is currently developed at NR and made available via Github. 
The package will provide a toolkit to evaluate predictions issued by ICPAC. `SeaVal` relies on R data tables (available with the `R` package `data.table`).
Data tables are more flexible and memory efficient data frames, and simplify many operations that are frequently required when working with weather- and climate data. An introduction to data tables can be found here: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
`SeaVal` will allow to convert the output of forecast models and observations generated at ICPAC into data tables. Moreover it will contain functionality for generating a variety of diagnostic plots, and provide various tools for forecast evaluation.


## Setting up `SeaVal`

In order to get started with the package you first need to install the packages `devtools` (if you don't have it already), which allows installing packages from Github directly:
```{r eval = FALSE, echo = TRUE}
install.packages('devtools')
```
Now you should be able to install the `SeaVal`-package like this:
```{r, eval = FALSE, echo = TRUE}
devtools::install_github('SeasonalForecastingEngine/SeaVal')
```

This line should not only install the `SeaVal` package, but also all packages it depends on. This may take a while, especially if you haven't installed some of the larger packages that we utilize, such as `data.table` or `ggplot2`.
Also, it is possible that you'll get a message like this:
```
These packages have more recent versions available.
It is recommended to update all of them.
Which would you like to update?

1: All                                 
2: CRAN packages only                  
3: None                             
...
```

In that case just type '1' for all. 

If this completes without an error, the setup is complete and you're good to go.
From here on out, all you have to do is load `SeaVal`:

```{r}
library(SeaVal)
```

This imports all functionality of the SeaVal package. It also imports the package `data.table`, so you don't need to put a separate `library(data.table)` in order to have data table syntax and functionality available.

## Handling data

Here we give a short introduction how to use common operations on data tables. The `SeaVal` package comes with a few example data sets, for example this one:

```{r}
data("chirps_monthly")
print(chirps_monthly)

```

We can look at a short description of the dataset like this:

```{r}
?chirps_monthly
```

We'll now go over a few basic commands for handling this sort of data. If we want to have a look at the OND-climatology of precipitation we can simply do this:

```{r}
clim_dt = chirps_monthly[,mean(prec),by = .(lon,lat)]
print(clim_dt)
ggplot_dt(clim_dt)
```



